---
title: "Creating CUPED from FWL (under construction)"
author: "Jeffrey Wong"
date: "2026-02-02"
categories: [experimentation, statistics]
---

```{r setup, include=FALSE}
require(dplyr)
require(purrr)
require(ggplot2)
require(gridExtra)
```

[CUPED](https://dl.acm.org/doi/pdf/10.1145/2433396.2433413) is a well known technique to greatly increase statistical power in a controlled experiment. The CUPED formulation has similarities to the OLS model

$$y = \beta_0 + T\beta_1 + (y_{\text{pre}}-1_n \mu_c) \beta_2 + \varepsilon$$

This blog post describes how the CUPED formulation can come out of the
[Frisch Waugh Lovell theorem](https://en.wikipedia.org/wiki/Frisch–Waugh–Lovell_theorem) for
ordinary least squares. This is my favorite way to derive the CUPED estimator,
and this exercise can help to build a deeper appreciation for the algebra of
variance reduction. I particularly like this formulation because the FWL is very
similar to the formulation of other modern causal ML techniques for variance reduction
like the R learner, MLRate, Double ML etc.

In this model, variation in $y$ is explained by the previous values contained in $y_{\text{pre}}$. For example, if someone watched a lot of movies during the pretreatment window, they are likely to watch a lot of movies during the posttreatment window too. Their particular $y$ values are explainable, and we can reduce the residual variance. This increases power.

While there are 3 parameters in the model, only 2 are meaningful. $\hat{\beta}_1$ represents the average treatment effect, and $\hat{\beta}_0$ represents the baseline value for the average user who is in the control, so $\hat{\beta}_0 = E[y | T = 0, y_{\text{pre}} = \mu_c]$. While $\beta_2$ is useful for decreasing variance, it is not useful for inference. It does not play a role in computing the treatment effect. $\beta_2$ is referred to as a nuisance variable.

The CUPED formulation has a special computing strategy, making it not only great for statistical power, but also easy to implement.

# The FWL

If a regression can be partitioned into variables of interest, $T$, and nuisance variables, $X$, then inference on $T$ can be done extremely efficiently.

$$y = \beta_0 + T\beta_1 + X\beta_2 + \varepsilon$$

1. Regress $y$ on $[1, X]$. Then get the residuals, $\tilde{y}$.
2. Regress $T$ on $[1, X]$. Then get the residuals, $\tilde{T}$.
3. Regress $\tilde{y}$ on $\tilde{T}$. The estimated parameters from this regression are exactly $\hat{\beta}_1$ from the original regression.

## Step 1

The regression $y$ on $[1, X]$ is a simple least squares regression with an intercept, e.g. $y = \gamma_{y, 0} + X\gamma_{y, 1} + \nu$. The solution for $\gamma$, and the residuals $\tilde{y}$ are simply

\begin{align}
\hat{\gamma}_{y, 1} &= \frac{Cov(X, y)}{Var(X)} \\
\hat{\gamma}_{y, 0} &= \bar{y} - \hat{\gamma}_{y, 1} \bar{X} \\
\tilde{T} &= y - (\hat{\gamma}_{y, 0} + X \hat{\gamma}_{y, 1})
\end{align}

## Step 2

The regression $T$ on $[1, X]$ can arguably be a no-op. $T$ is by definition not correlated with anything, so the slope of this regression must be 0. The intercept
would just be the proportion of users that are in the treatment. This elegant
subtlety is leveraged in CUPED.

\begin{align}
\hat{\gamma}_{T, 1} &= 0 \\
\hat{\gamma}_{T, 0} &= \bar{T} \\
\tilde{T} &= y - (\bar{T})
\end{align}

However, a fully complete OLS solution would say

\begin{align}
\hat{\gamma}_{T, 1} &= \frac{Cov(X, T)}{Var(X)} \\
\hat{\gamma}_{T, 0} &= \bar{T} - \hat{\gamma}_{T, 1} \bar{X} \\
\tilde{T} &= T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})
\end{align}

## Step 3

Now we regress $\tilde{y}$ on $\tilde{T}$ to get the treatment effect.

$$\hat{\beta} = \frac{Cov(\tilde{y}, \tilde{T})}{Var(\tilde{T})}$$

The numerator can be consolidated.
\begin{align}
Cov(\tilde{y}, \tilde{T}) &= Cov(y - (\hat{\gamma}_{y, 0} + X \hat{\gamma}_{y, 1}), T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})) \\
  &= Cov(y, T) - \hat{\gamma}_{T, 1} Cov(y, X) - \hat{\gamma}_{y, 1} Cov(X, T) + \hat{\gamma}_{y, 1} \hat{\gamma}_{T, 1} Var(X)
\end{align}

The second and fourth terms cancel out because $\hat{\gamma}_{y, 1} = \frac{Cov(X, y)}{Var(X)}$, leaving
$$Cov(\tilde{y}, \tilde{T}) = Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)$$

The denominator is also consolidated
\begin{align}
Var(\tilde{T}) &= Var(T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})) \\
  &= Var(T) - 2\hat{\gamma}_{T, 1} Cov(T, X) + \hat{\gamma}_{T, 1}^2 Var(X)
\end{align}

Note $\hat{\gamma}_{T, 1} = \frac{Cov(T, X)}{Var(X)}$, which buys the reduction
$$Var(\tilde{T}) = Var(T) - \hat{\gamma}_{T, 1} Cov(T, X)$$
This yields

\begin{align}
\hat{\gamma}_{y, 1} &= \frac{Cov(X, y)}{Var(X)} \\
\hat{\gamma}_{T, 1} &= \frac{Cov(X, T)}{Var(X)} \\
\hat{\beta} &= \frac{Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)}{Var(T) - \hat{\gamma}_{T, 1} Cov(X, T)}
\end{align}

In the same spirit of [conditionally sufficient statistics](https://arxiv.org/pdf/2102.11297),
this compute strategy for the average treatment effect, controlling for a 
continuous covariate, can be estimated entirely with aggregated data.

## Estimating $\hat{\beta}_2$

Our original linear model is

$$y = \beta_0 + T\beta_1 + X\beta_2 + \varepsilon.$$

Thinking of $\beta_2$ as a nuisance variable, we partialed it out using the FWL.
However, there is a way to compute $(\hat{\beta}_1, \hat{\beta}_2)$ simultaneously.
Having $(\hat{\beta}_1, \hat{\beta}_2)$ will be useful for computing robust
standard errors below.

Consider the demeaned version of the problem, where there is no intercept.

$$y = T\beta_1 + X\beta_2 + \varepsilon.$$

Take the covariance with respect to $T$ on both sides to get

\begin{align}
Cov(y, T) &= Cov(T\beta_1 + X\beta_2 + \varepsilon, T) \\
  &= \beta_1 Var(T) + \beta_2 Cov(X, T) + Cov(\varepsilon, T) \\
  &= \beta_1 Var(T) + \beta_2 Cov(X, T)
\end{align}

where, at the least squares solution, $Cov(\varepsilon, T) = 0$.
Similarly

\begin{align}
Cov(y, X) &= Cov(T\beta_1 + X\beta_2 + \varepsilon, X) \\
  &= \beta_1 Cov(T, X) + \beta_2 Var(X) + Cov(\varepsilon, X) \\
  &= \beta_1 Cov(T, X) + \beta_2 Var(X)
\end{align}

Because of this property of covariances, we can formulate a system
relating $(\beta_1, \beta_2)$

$$
\begin{bmatrix} 
Var(T) & Cov(T, X) \\ 
Cov(T, X) & Var(X) 
\end{bmatrix} 
\begin{bmatrix} 
\hat{\beta}_1 \\ 
\hat{\beta}_2
\end{bmatrix} 
= 
\begin{bmatrix} 
Cov(y, T) \\ 
Cov(y, X) 
\end{bmatrix}
$$

This is a tiny $2 \times 2$ system. Using [Cramer's Rule](https://en.wikipedia.org/wiki/Cramer%27s_rule),
we solve for the coefficients. First, we define the determinant of the covariance matrix.
Then the estimates are

\begin{align}
D &= Var(T)Var(X) - Cov(T, X)^2 \\
\hat{\beta}_1 &= \frac{Var(X)Cov(y, T) - Cov(T, X)Cov(y, X)}{D} \\
\hat{\beta}_2 &= \frac{Var(T)Cov(y, X) - Cov(T, X)Cov(y, T)}{D}
\end{align}

## Standard Error for the Average Treatment Effect

If we were fitting OLS, we would care about fitting robust standard errors
for the parameters of the model. Using raw data from users, that can be done by
simply fitting $\tilde{y}$ on $\tilde{T}$ and invoking standard errors.
To fit the standard errors using aggregated data alone, we would need to keep track of
the first, second, third, and fourth moments of the features. This is quite a hassle.

Instead, we are going to relax what we mean by "robust". Under standard OLS,
we would say that the variance for an individual residual, $\varepsilon_i$ is a 
function of that individual's feature vector. We won't need this granularity
given that we only care about the standard error of the average of the treatment effect.

Again, in the same spirit of [conditionally sufficient statistics](https://arxiv.org/pdf/2102.11297),
say there are $G$ groups of users in the data. For simplicity, say $G = 2$ and
the groups are simply the control group and treatment group. We seek a group robust
standard error, meaning the variance of the residuals for the control group
can be different than the variance of the residuals for the treatment group.
This leads us to

$$\varepsilon_{i, g} = y_{i,g} - (\hat{\beta}_0 + \hat{\beta}_1 T_i + \hat{\beta}_2 y_{i, \text{pre}})$$
If $g$ is defined by treatment or control, then $\hat{\beta}_0$ and $\hat{\beta}_1 T_i$ are
constants within group $g$. Thus, the group variance is simply
$$\sigma^2_g = Var(y_g - \hat{\beta}_2 y_{\text{pre}, g}) = Var(y_g) + \hat{\beta}_2^2 Var(y_{\text{pre}, g}) - 2 Cov(y_g, y_{\text{pre}, g})$$
Using this formula for group variance we can compute $\sigma^2_T$ and $\sigma^2_C$.
We can express the general formula for the variance of the ATE in terms of these variables.

$$Var(\widehat{ATE}) = Var(\hat{\beta}_1) = \sigma^2_T/n_T + \sigma^2_C/n_C.$$

# CUPED is a special case.

CUPED leverages *perfectly* randomized data, so $Cov(X, T) = 0$. Now we will rewrite
the solution for $\hat{\beta}$ in a clever way.

\begin{align}
\hat{\beta} &= \frac{Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)}{Var(T)} \\
  &= \frac{Cov(y, T)}{Var(T)} - \hat{\gamma}_{y, 1} \frac{Cov(X, T)}{Var(T)}
\end{align}
which looks like the difference between 2 different OLS solutions: $y$ regressed on $T$
and $X$ regressed on $T$. Knowing $T$ is binary, each of these OLS solutions is simply
the difference in group means:

$$\hat{\beta} = E[y | T = 1] - E[y | T = 0] - \hat{\gamma}_{y, 1} (E[X | T = 1] - E[X | T = 0]).$$
This is exactly the CUPED formulation! However, it truly reveals that CUPED
is like FWL with an assumption that $E[X | T = 1] = E[X | T = 0]$. In small
sample bias this may not be the case; by luck a user with large values for $X$
can be randomized into the treatment. In these scenarios, a pretreatment bias
may be triggered. If it was due to randomization, instead of selection bias, we
would simply control for it by modeling it into the outcome equation. Enforcing
these values to be equal does not let us control for different baselines per arm.
We lean into adopting the full FWL solution, since it also has an efficient compute strategy.

Nonetheless, to round out this section we derive robust standard errors for the ATE
for CUPED. By enforcing the property $Cov(X, T) = 0$, we can efficiently estimate
$\hat{\beta}_2 = \frac{Cov(y, X)}{Var(X)}$. This pops out by using FWL from the other
direction, $\beta_2$ is the parameter of interest and we want to partial out the
effect from $T$. Once we have $\hat{\beta}_2$, the exercise for the standard
error of CUPED is the same as above.

# For Relative Effects

The relative effect wants to compute $$R = \frac{\hat{\beta}_1}{E[y | T = 0, y_\text{pre} = \bar{y}_\text{pre}]} = \frac{\hat{\beta}_1}{\hat{\beta}_0 + \hat{\beta}_2 \bar{y}_\text{pre}}$$

We are going to take advantage of a clever rewrite of the denominator.
In general, the intercept of a regression is computed as $\hat{\beta}_0 = \bar{y}_C - \hat{\beta}_2 \bar{X}_C$. Then, we can make the rewrite

\begin{align}
\hat{\mu}_C &= E[y | T = 0, y_\text{pre} = \bar{y}_\text{pre}] \\
  &= \hat{\beta}_0 + \hat{\beta}_2 \bar{y}_\text{pre} \\
  &= \bar{y}_C - \hat{\beta}_2 (\bar{X}_C - \bar{X}) \\
R &= \frac{\hat{\beta}_1}{\hat{\mu}_C}
\end{align}
The relative effect can be computed using aggregated data.

The standard error is computed using the delta method. In general, when the relative
effect is expressed as the ratio of two random variables, $R = \frac{N}{D}$, the standard error
of the estimator is 
$$Var\left( \frac{\hat{N}}{\hat{D}} \right) \approx \frac{Var(\hat{N})}{\mu_D^2} + \frac{\mu_N^2 Var(\hat{D})}{\mu_D^4} - \frac{2 \mu_N Cov(\hat{N}, \hat{D})}{\mu_D^3}$$

In the case of the relative effect, we will need to define the variance of
the denominator.

\begin{align}
Var(\hat{\mu}_C) &= Var(\bar{y}_C) + Var(\hat{\beta}_2 (\bar{X}_C - \bar{X})) - 2Cov(\bar{y}_C, \hat{\beta}_2 (\bar{X}_C - \bar{X})) \\
  &= \sigma^2_0 / n_0 + (\bar{X}_C - \bar{X})^2 Var(\hat{\beta}_2)
\end{align}
Because of centering, the covariance term is zero. 
The final variance of the relative effect uses

\begin{align}
\sigma^2 &= \frac{(n_T - 1) \sigma^2_T + (n_C - 1) \sigma^2_C}{N - 3} \\
SS_X &= (N-1) Var(X) \\
R^2_{X,T} &= \frac{Cov(T, X)^2}{Var(T)Var(X)} \\
Var(\hat{\beta}_2) &= \frac{\sigma^2}{SS_X (1 - R^2_{X,T})} \\
Var(\hat{\mu}_C) &= \sigma^2_0 / n_0 + (\bar{X}_C - \bar{X})^2 Var(\hat{\beta}_2) \\
Var(\hat{R}) &\approx \frac{Var(\hat{\beta}_1)}{\hat{\mu}_C^2} + 
  \frac{\hat{\beta}_1^2 Var(\hat{\mu}_C)}{\hat{\mu}_C^4} - 
  \frac{2 \hat{\beta}_1 Cov(\hat{\beta}_1, \hat{\mu}_C)}{\hat{\mu}_C^3}
\end{align}

Under perfect randomization $\bar{X}_C = \bar{X}_T = \bar{X}$ so $\hat{\mu}_C = \sigma^2_0 / n_0$,
removing the need to track $Var(\hat{\beta}_2)$. This is a very convenient property.

# Subtle differences between CUPED and OLS

CUPED is very similar to OLS, but it is not the same. Throughout this post,
we showed that CUPED is actually a special case of OLS, where it is assumed that
the experiment is perfectly balanced, so $\bar{X}_T = \bar{X}_C = \bar{X}$,
and that there is no pretreatment bias, so $Cov(T, X) = 0$. These assumptions
were used in multiple places.

1. It was used in the estimation of the average effect by skipping step 2 of FWL.
2. It was used in the standard error of the ATE.
3. It was used to compute $\hat{\mu}_C$, the mean of the adjusted baseline,
which was subsequently used to compute the variance of the relative effect.

# Code

```{r}
# Function to compute treatment effects from aggregated data
compute_ab_stats <- function(n1, sum_y1, sum_p1, sum_y2_1, sum_p2_1, sum_yp1,
                             n0, sum_y0, sum_p0, sum_y2_0, sum_p2_0, sum_yp0) {
  
  # 1. Combined Aggregates
  n <- n1 + n0
  s_y <- sum_y1 + sum_y0
  s_p <- sum_p1 + sum_p0
  s_yp <- sum_yp1 + sum_yp0
  s_p2 <- sum_p2_1 + sum_p2_0
  
  # Means
  y_bar1 <- sum_y1 / n1
  y_bar0 <- sum_y0 / n0
  p_bar1 <- sum_p1 / n1
  p_bar0 <- sum_p0 / n0
  p_bar_total <- s_p / n
  t_bar <- n1 / n
  
  # 2. Covariances (Global)
  var_p <- (s_p2 / n) - (p_bar_total^2)
  cov_yp <- (s_yp / n) - (s_y / n) * (s_p / n)
  cov_tp <- (sum_p1 / n) - (t_bar * p_bar_total)
  cov_ty <- (sum_y1 / n) - (t_bar * (s_y / n))
  var_t <- t_bar * (1 - t_bar)
  
  # 3. FWL Coefficients
  gamma_y <- cov_yp / var_p
  gamma_t <- cov_tp / var_p
  
  # ATE (Beta) - Adjusted for noisy baseline
  ate <- (cov_ty - gamma_y * cov_tp) / (var_t - gamma_t * cov_tp)
  
  # 4. Group-Robust Variances
  # Var(y - gamma*p) = Var(y) + gamma^2*Var(p) - 2*gamma*Cov(y,p)
  var_y1 <- (sum_y2_1 / n1) - (y_bar1^2)
  var_p1 <- (sum_p2_1 / n1) - (p_bar1^2)
  cov_yp1 <- (sum_yp1 / n1) - (y_bar1 * p_bar1)
  resid_var1 <- var_y1 + (gamma_y^2 * var_p1) - (2 * gamma_y * cov_yp1)
  
  var_y0 <- (sum_y2_0 / n0) - (y_bar0^2)
  var_p0 <- (sum_p2_0 / n0) - (p_bar0^2)
  cov_yp0 <- (sum_yp0 / n0) - (y_bar0 * p_bar0)
  resid_var0 <- var_y0 + (gamma_y^2 * var_p0) - (2 * gamma_y * cov_yp0)
  
  # Standard Error of ATE
  se_ate <- sqrt((resid_var1 / n1) + (resid_var0 / n0))
  
  # 5. Relative Effect (Delta Method)
  mu0_adj <- y_bar0 - gamma_y * (p_bar0 - p_bar_total)
  rel_effect <- ate / mu0_adj
  
  # SE of Relative Effect (Delta Method approx)
  # Var(A/B) approx (Var(A)/B^2) + (A^2 * Var(B)/B^4)
  var_mu0 <- resid_var0 / n0
  se_rel <- sqrt((se_ate^2 / mu0_adj^2) + (ate^2 * var_mu0 / mu0_adj^4))
  
  return(list(
    ate = ate,
    se_ate = se_ate,
    relative_effect = rel_effect,
    se_relative = se_rel,
    p_value = 2 * (1 - pnorm(abs(ate / se_ate)))
  ))
}
```