---
title: "Creating CUPED from FWL (under construction)"
author: "Jeffrey Wong"
date: "2026-02-02"
categories: [experimentation, statistics]
---

```{r setup, include=FALSE}
require(dplyr)
require(purrr)
require(ggplot2)
require(gridExtra)
require(sandwich)
require(lmtest)
```

[CUPED](https://dl.acm.org/doi/pdf/10.1145/2433396.2433413) is a well known technique to greatly increase statistical power in a controlled experiment. The CUPED formulation has similarities to the OLS model

$$y = \beta_0 + T\beta_1 + (y_{\text{pre}}-1_n \mu_c) \beta_2 + \varepsilon$$

This blog post describes how the CUPED formulation can come out of the
[Frisch Waugh Lovell theorem](https://en.wikipedia.org/wiki/Frisch–Waugh–Lovell_theorem) for
ordinary least squares. This is my favorite way to derive the CUPED estimator,
and this exercise can help to build a deeper appreciation for the algebra of
variance reduction. I particularly like this formulation because the FWL is very
similar to the formulation of other modern causal ML techniques for variance reduction
like the R learner, MLRate, Double ML etc.

In this model, variation in $y$ is explained by the previous values contained in $y_{\text{pre}}$. For example, if someone watched a lot of movies during the pretreatment window, they are likely to watch a lot of movies during the posttreatment window too. Their particular $y$ values are explainable, and we can reduce the residual variance. This increases power.

While there are 3 parameters in the model, only 2 are meaningful. $\hat{\beta}_1$ represents the average treatment effect, and $\hat{\beta}_0$ represents the baseline value for the average user who is in the control, so $\hat{\beta}_0 = E[y | T = 0, y_{\text{pre}} = \mu_c]$. While $\beta_2$ is useful for decreasing variance, it is not useful for inference. It does not play a role in computing the treatment effect. $\beta_2$ is referred to as a nuisance variable.

The CUPED formulation has a special computing strategy, making it not only great for statistical power, but also easy to implement.

# The FWL

If a regression can be partitioned into variables of interest, $T$, and nuisance variables, $X$, then inference on $T$ can be done extremely efficiently.

$$y = \beta_0 + T\beta_1 + X\beta_2 + \varepsilon$$

1. Regress $y$ on $[1, X]$. Then get the residuals, $\tilde{y}$.
2. Regress $T$ on $[1, X]$. Then get the residuals, $\tilde{T}$.
3. Regress $\tilde{y}$ on $\tilde{T}$. The estimated parameters from this regression are exactly $\hat{\beta}_1$ from the original regression.

## Step 1

The regression $y$ on $[1, X]$ is a simple least squares regression with an intercept, e.g. $y = \gamma_{y, 0} + X\gamma_{y, 1} + \nu$. The solution for $\gamma$, and the residuals $\tilde{y}$ are simply

\begin{align}
\hat{\gamma}_{y, 1} &= \frac{Cov(X, y)}{Var(X)} \\
\hat{\gamma}_{y, 0} &= \bar{y} - \hat{\gamma}_{y, 1} \bar{X} \\
\tilde{T} &= y - (\hat{\gamma}_{y, 0} + X \hat{\gamma}_{y, 1})
\end{align}

## Step 2

The regression $T$ on $[1, X]$ can arguably be a no-op. $T$ is by definition not correlated with anything, so the slope of this regression must be 0. The intercept
would just be the proportion of users that are in the treatment. This elegant
subtlety is leveraged in CUPED.

\begin{align}
\hat{\gamma}_{T, 1} &= 0 \\
\hat{\gamma}_{T, 0} &= \bar{T} \\
\tilde{T} &= y - (\bar{T})
\end{align}

However, a fully complete OLS solution would say

\begin{align}
\hat{\gamma}_{T, 1} &= \frac{Cov(X, T)}{Var(X)} \\
\hat{\gamma}_{T, 0} &= \bar{T} - \hat{\gamma}_{T, 1} \bar{X} \\
\tilde{T} &= T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})
\end{align}

## Step 3

Now we regress $\tilde{y}$ on $\tilde{T}$ to get the treatment effect.

$$\hat{\beta}_1 = \frac{Cov(\tilde{y}, \tilde{T})}{Var(\tilde{T})}$$

The numerator can be consolidated.
\begin{align}
Cov(\tilde{y}, \tilde{T}) &= Cov(y - (\hat{\gamma}_{y, 0} + X \hat{\gamma}_{y, 1}), T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})) \\
  &= Cov(y, T) - \hat{\gamma}_{T, 1} Cov(y, X) - \hat{\gamma}_{y, 1} Cov(X, T) + \hat{\gamma}_{y, 1} \hat{\gamma}_{T, 1} Var(X)
\end{align}

The second and fourth terms cancel out because $\hat{\gamma}_{y, 1} = \frac{Cov(X, y)}{Var(X)}$, leaving
$$Cov(\tilde{y}, \tilde{T}) = Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)$$

The denominator is also consolidated
\begin{align}
Var(\tilde{T}) &= Var(T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})) \\
  &= Var(T) - 2\hat{\gamma}_{T, 1} Cov(T, X) + \hat{\gamma}_{T, 1}^2 Var(X)
\end{align}

Note $\hat{\gamma}_{T, 1} = \frac{Cov(T, X)}{Var(X)}$, which buys the reduction
$$Var(\tilde{T}) = Var(T) - \hat{\gamma}_{T, 1} Cov(T, X)$$
This yields

\begin{align}
\hat{\gamma}_{y, 1} &= \frac{Cov(X, y)}{Var(X)} \\
\hat{\gamma}_{T, 1} &= \frac{Cov(X, T)}{Var(X)} \\
\hat{\beta}_1 &= \frac{Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)}{Var(T) - \hat{\gamma}_{T, 1} Cov(X, T)}
\end{align}

In the same spirit of [conditionally sufficient statistics](https://arxiv.org/pdf/2102.11297),
this compute strategy for the average treatment effect, controlling for a 
continuous covariate, can be estimated entirely with aggregated data.

## Estimating $\hat{\beta}_2$

Our original linear model is

$$y = \beta_0 + T\beta_1 + X\beta_2 + \varepsilon.$$

Thinking of $\beta_2$ as a nuisance variable, we partialed it out using the FWL.
However, there is a way to compute $(\hat{\beta}_1, \hat{\beta}_2)$ simultaneously.
Having $(\hat{\beta}_1, \hat{\beta}_2)$ will be useful for computing robust
standard errors below.

Consider the demeaned version of the problem, where there is no intercept.

$$y = T\beta_1 + X\beta_2 + \varepsilon.$$

Take the covariance with respect to $T$ on both sides to get

\begin{align}
Cov(y, T) &= Cov(T\beta_1 + X\beta_2 + \varepsilon, T) \\
  &= \beta_1 Var(T) + \beta_2 Cov(X, T) + Cov(\varepsilon, T) \\
  &= \beta_1 Var(T) + \beta_2 Cov(X, T)
\end{align}

where, at the least squares solution, $Cov(\varepsilon, T) = 0$.
Similarly

\begin{align}
Cov(y, X) &= Cov(T\beta_1 + X\beta_2 + \varepsilon, X) \\
  &= \beta_1 Cov(T, X) + \beta_2 Var(X) + Cov(\varepsilon, X) \\
  &= \beta_1 Cov(T, X) + \beta_2 Var(X)
\end{align}

Because of this property of covariances, we can formulate a system
relating $(\beta_1, \beta_2)$

$$
\begin{bmatrix} 
Var(T) & Cov(T, X) \\ 
Cov(T, X) & Var(X) 
\end{bmatrix} 
\begin{bmatrix} 
\beta_1 \\ 
\beta_2
\end{bmatrix} 
= 
\begin{bmatrix} 
Cov(y, T) \\ 
Cov(y, X) 
\end{bmatrix}
$$

This is a tiny $2 \times 2$ system. Using [Cramer's Rule](https://en.wikipedia.org/wiki/Cramer%27s_rule),
we write a solution for the coefficients in closed form. 
First, we define the determinant of the covariance matrix. Then the estimates are

\begin{align}
D &= Var(T)Var(X) - Cov(T, X)^2 \\
\hat{\beta}_1 &= \frac{Var(X)Cov(y, T) - Cov(T, X)Cov(y, X)}{D} \\
\hat{\beta}_2 &= \frac{Var(T)Cov(y, X) - Cov(T, X)Cov(y, T)}{D}
\end{align}

# Heteroskedastic Consistent Covariances (Robust Standard Errors)

The above section covers how to fit point estimates. Below we will discuss the variance
of the effect, approximations, and how to estimate the variance using only
aggregates of data.

Normally in OLS we would want to fit the parameters using robust standard errors.
This says that not only is the expectation of $y$ a linear combination of the
features and the parameters, the variance is also a function of the parameters. 
We see this in the history of t tests already, though t tests don't make use of $X$.
When computing the standard error for the average effect according to a t test, we calculate

$$Var(\widehat{ATE}) = \hat{\sigma}^2_T / n_T + \hat{\sigma}^2_C / n_C$$
which fits separate variance terms for the treatment and the control. In an OLS
model, the t test is the same as fitting $y = \beta_0 + T\beta_1 + \varepsilon$ with
[HC0 robust standard errors](https://en.wikipedia.org/wiki/Heteroskedasticity-consistent_standard_errors).
When there is only 1 feature, we are saying the
variance of $y$ is allowed to vary, but it simply varies across the
$T = 0$ group and the $T = 1$ group. The HC0 robust standard error in this scenario
could have also been called a "group robust standard error".

As we introduce $X$ into the regression, the OLS formulation would say
the variance of each $y_i$ is different and is a function of the $x_i$.
This is formalized through the sandwich estimator.

\begin{align}
M &= \begin{bmatrix} 1 & T & X \end{bmatrix} \\
Var(\hat{\beta}) &= (M^T M)^{-1} \left( \sum_{i=1}^{n} m_i m_i^T \hat{\varepsilon}_i^2 \right) (M^T M)^{-1}
\end{align}

After running FWL steps 1-3, we can invoke the sandwich estimator using the residuals
$\tilde{y}$ and $\tilde{T}$. 

## Compute Strategy with 4th order moments

The point estimate for the parameters can be derived from just means, variances,
and covariances. This applies to both FWL and CUPED.
To compute the traditional sandwich, we either need the individual level data,
or we need fourth order aggregates. The bread matrix can be computed using
second order aggregates, which is what is used to get the point estimates.
However, the meat matrix is much more challenging. The intuition behind it is that
the $\varepsilon$ term has $x$ in it, so $m_i m_i^T \hat{\varepsilon}_i^2$ has
$x^4$ in it. If we are able to store the 4th order moments, we can completely recover the HC0
covariance matrix. But this is a hassle. If we want to limit ourselves to the best approximation to the
standard error using just second order aggregates, we can formulate the
group robust standard error.

## Heteroskedastic Consistent (by groups) with only 2nd order moments

Finding an estimator that only uses 2nd order moments would be very convenient.
The best estimator operating within these constraints is the HC by group sandwich
estimator. Similar to how the t test allowed variability across the treatment
and control groups, we allow the meat matrix to vary by groups. However, we do not
allow the meat matrix to vary by individual $m_i$ and $\hat{\varepsilon}_i$.
The formulation would be 
$$\widehat{\text{Var}}(\hat{\beta}) = (M^T M)^{-1} \left[ \sum_g \sigma_g^2 (X_g^T X_g) \right] (M^T M)^{-1}$$

The residual variance for a group $g$ is $\sigma^2_g$, and is defined
$$Var(\varepsilon_{i, g}) = Var(y_{i,g} - (\hat{\beta}_0 + \hat{\beta}_1 T_i + \hat{\beta}_2 y_{i, \text{pre}}))$$

If $g$ is defined by treatment or control, then $\hat{\beta}_0$ and $\hat{\beta}_1 T_i$ are
constants within group $g$. Thus, the group variance is simply
$$\sigma^2_g = Var(y_g - \hat{\beta}_2 y_{\text{pre}, g}) = Var(y_g) + \hat{\beta}_2^2 Var(y_{\text{pre}, g}) - 2 Cov(y_g, y_{\text{pre}, g})$$

The entire covariance matrix for the parameters can be derived from aggregated data.
Remember, this is the best approximation possible to the sandwich when only 2nd order
information is available.

# CUPED is a special case.

CUPED leverages *perfectly* randomized data, so $Cov(X, T) = 0$. This effectively
makes step 2 of FWL a no-op. Now we will rewrite
the solution for $\hat{\beta}_1$ in a clever way, based on our FWL solution.

\begin{align}
\hat{\beta}_1 &= \frac{Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)}{Var(T)} \\
  &= \frac{Cov(y, T)}{Var(T)} - \hat{\gamma}_{y, 1} \frac{Cov(X, T)}{Var(T)}
\end{align}
which looks like the difference between 2 different OLS solutions: $y$ regressed on $T$
and $X$ regressed on $T$. Knowing $T$ is binary, each of these OLS solutions is simply
the difference in group means:

\begin{align}
\hat{\beta}_1 &= E[y | T = 1] - E[y | T = 0] - \hat{\gamma}_{y, 1} (E[X | T = 1] - E[X | T = 0]) \\
  &= \biggl(E[y | T = 1] - \hat{\gamma}_{y, 1} E[X | T = 1] \biggr) - \biggl(E[y | T = 0] - \hat{\gamma}_{y, 1} E[X | T = 0] \biggr)
\end{align}

**This is exactly the CUPED formulation!** It is a difference in means
that adjusts for a difference in the pretreatment baseline. Or, it can be thought
of as the difference in the means of the adjusted $y$ variable.
CUPED is an output of FWL when step 2 is skipped. But this requires $Cov(X, T) = 0$. 

Since CUPED is a difference in the means of the adjusted $y$, its variance looks
just like the variance of a t test.
$$Var(\widehat{ATE}) = \frac{Var(y_T - \hat{\gamma}_{y, 1}x_T)}{n_T} + \frac{Var(y_C - \hat{\gamma}_{y, 1} x_C)}{n_C}.$$

Ironically, this formulation of the variance is neither equal to HC variances,
nor is it equal to group robust variances. This variance simply weights the variance
of the residuals according to their sample size. The group robust variance weights
the residuals according to $X_g^T X_g$. Under CUPED, $\hat{\gamma}_{y, 1}$ is assumed to
have no estimation variance of its own, it is a fixed constant. Moreover, that
$\hat{\gamma}_{y, 1}$ variable is derived using the assumption $Cov(X, T) = 0$.

CUPED and OLS (via 2nd order aggs) differ for small samples, and cases of pretreatment bias.
By luck, a user with large values for $X$ can be randomized into the treatment. 
In these scenarios, a pretreatment bias may be triggered and CUPED's assumption
of $Cov(T, X) = 0$ is vulnerable. On the other hand, OLS will remain robust.

# For Relative Effects

The relative effect wants to compute $$R = \frac{\hat{\beta}_1}{E[y | T = 0, y_\text{pre} = \bar{y}_\text{pre}]} = \frac{\hat{\beta}_1}{\hat{\beta}_0 + \hat{\beta}_2 \bar{y}_\text{pre}}$$

We are going to take advantage of a clever rewrite of the denominator.
In general, the intercept of a regression is computed as $\hat{\beta}_0 = \bar{y}_C - \hat{\beta}_2 \bar{X}_C$. Then, we can make the rewrite

\begin{align}
\hat{\mu}_C &= E[y | T = 0, y_\text{pre} = \bar{y}_\text{pre}] \\
  &= \hat{\beta}_0 + \hat{\beta}_2 \bar{y}_\text{pre} \\
  &= \bar{y}_C - \hat{\beta}_2 (\bar{X}_C - \bar{X}) \\
R &= \frac{\hat{\beta}_1}{\hat{\mu}_C}
\end{align}
The relative effect can be computed using aggregated data.

The standard error is computed using the delta method. In general, when the relative
effect is expressed as the ratio of two random variables, $R = \frac{N}{D}$, the standard error
of the estimator is 
$$Var\left( \frac{\hat{N}}{\hat{D}} \right) \approx \frac{Var(\hat{N})}{\mu_D^2} + \frac{\mu_N^2 Var(\hat{D})}{\mu_D^4} - \frac{2 \mu_N Cov(\hat{N}, \hat{D})}{\mu_D^3}$$

In the case of the relative effect, we will need to define the variance of
the denominator.

\begin{align}
Var(\hat{\mu}_C) &= Var(\bar{y}_C) + Var(\hat{\beta}_2 (\bar{X}_C - \bar{X})) - 2Cov(\bar{y}_C, \hat{\beta}_2 (\bar{X}_C - \bar{X})) \\
  &= \sigma^2_0 / n_0 + (\bar{X}_C - \bar{X})^2 Var(\hat{\beta}_2)
\end{align}
Because of centering, the covariance term is zero. 
The final variance of the relative effect uses

\begin{align}
\sigma^2 &= \frac{(n_T - 1) \sigma^2_T + (n_C - 1) \sigma^2_C}{N - 3} \\
SS_X &= (N-1) Var(X) \\
R^2_{X,T} &= \frac{Cov(T, X)^2}{Var(T)Var(X)} \\
Var(\hat{\beta}_2) &= \frac{\sigma^2}{SS_X (1 - R^2_{X,T})} \\
Var(\hat{\mu}_C) &= \sigma^2_0 / n_0 + (\bar{X}_C - \bar{X})^2 Var(\hat{\beta}_2) \\
Var(\hat{R}) &\approx \frac{Var(\hat{\beta}_1)}{\hat{\mu}_C^2} + 
  \frac{\hat{\beta}_1^2 Var(\hat{\mu}_C)}{\hat{\mu}_C^4} - 
  \frac{2 \hat{\beta}_1 Cov(\hat{\beta}_1, \hat{\mu}_C)}{\hat{\mu}_C^3}
\end{align}

Under perfect randomization $\bar{X}_C = \bar{X}_T = \bar{X}$ so $\hat{\mu}_C = \sigma^2_0 / n_0$,
removing the need to track $Var(\hat{\beta}_2)$. This is a very convenient property.

# Code

```{r}
set.seed(100)
n = 1000
data = data.frame(
  treated = sample(0:1, n, replace = TRUE),
  baseline = rnorm(n),
  x = rnorm(n),
  effect_if_treated = rnorm(n, 2, 1)
) %>%
  mutate(
    effect = treated * effect_if_treated,
    y = baseline + effect + 0.5 * x
  )
model = lm(y ~ treated + x, data)

coeftest(model, vcovHC(model, "HC0"))
```

```{r}
#' Compute Treatment Effects from Aggregated Data using FWL
#'
#' @description Calculates the Average Treatment Effect (ATE) and Relative Lift 
#' using the Frisch-Waugh-Lovell (FWL) theorem. This approach effectively 
#' residualizes the outcome and treatment against the covariate using 
#' aggregate-level sufficient statistics.
#'
#' @param n_treatment Count of users in Treatment
#' @param sum_y_treatment Sum of outcomes in Treatment
#' @param sum_x_treatment Sum of covariates in Treatment
#' @param sum_y2_treatment Sum of squared outcomes in Treatment
#' @param sum_x2_treatment Sum of squared covariates in Treatment
#' @param sum_y_x_treatment Sum of cross-products (y*p) in Treatment
#' @param n_control Count of users in Control
#' @param sum_y_control Sum of outcomes in Control
#' @param sum_x_control Sum of covariates in Control
#' @param sum_y2_control Sum of squared outcomes in Control
#' @param sum_x2_control Sum of squared covariates in Control
#' @param sum_y_x_control Sum of cross-products (y*p) in Control
#'
#' @return A list containing:
#' \itemize{
#'   \item \code{ate}: The absolute average treatment effect.
#'   \item \code{se_ate}: The robust standard error of the ATE.
#'   \item \code{relative_effect}: The lift relative to the control baseline.
#'   \item \code{se_relative}: The standard error of the relative effect.
#'   \item \code{p_value}: Two-tailed p-value for the ATE.
#' }
#' 
#' @export
compute_ab_stats <- function(n_treatment, sum_y_treatment, sum_x_treatment, sum_y2_treatment, sum_x2_treatment, sum_y_x_treatment,
                             n_control, sum_y_control, sum_x_control, sum_y2_control, sum_x2_control, sum_y_x_control) {

  # Sums
  n <- n_treatment + n_control
  sum_y <- sum_y_treatment + sum_y_control
  sum_x <- sum_x_treatment + sum_x_control
  sum_y_x <- sum_y_x_treatment + sum_y_x_control
  sum_x2 <- sum_x2_treatment + sum_x2_control
  
  # Means
  mean_y_treatment <- sum_y_treatment / n_treatment
  mean_y_control <- sum_y_control / n_control
  mean_x_treatment <- sum_x_treatment / n_treatment
  mean_x_control <- sum_x_control / n_control
  mean_x <- sum_x / n
  mean_t <- n_treatment / n
  
  # Covariances
  var_x <- ((sum_x2 / n) - (mean_x^2)) * (n / (n-1))
  cov_y_x <- ((sum_y_x / n) - (sum_y / n) * (sum_x / n)) * (n / (n-1))
  cov_t_x <- ((sum_x_treatment / n) - (mean_t * mean_x)) * (n / (n-1))
  cov_t_y <- ((sum_y_treatment / n) - (mean_t * (sum_y / n))) * (n / (n-1))
  var_t <- mean_t * (1 - mean_t) # bernoulli variance

  # FWL Coefficients
  gamma_y <- cov_y_x / var_x
  gamma_t <- cov_t_x / var_x

  # Group-Robust Variances
  # Residual variance calculation (Var(y - gamma*x))
  var_from_aggs = function(sum_y2, n, mean_y) {
    ((sum_y2 / n) - (mean_y)^2) * (n / (n - 0))
  }
  
  var_y_treatment = var_from_aggs(sum_y2_treatment, n_treatment, mean_y_treatment)
  var_x_treatment = var_from_aggs(sum_x2_treatment, n_treatment, mean_x_treatment)
  cov_y_x_treatment <- ((sum_y_x_treatment / n_treatment) - (mean_y_treatment * mean_x_treatment))  * (n_treatment / (n_treatment-0))
  resid_var1 <- var_y_treatment + (gamma_y^2 * var_x_treatment) - (2 * gamma_y * cov_y_x_treatment)

  var_y_control <- var_from_aggs(sum_y2_control, n_control, mean_y_control)
  var_x_control <- var_from_aggs(sum_x2_control, n_control, mean_x_control)
  cov_y_x_control <- ((sum_y_x_control / n_control) - (mean_y_control * mean_x_control)) * (n_control / (n_control-0))
  resid_var0 <- var_y_control + (gamma_y^2 * var_x_control) - (2 * gamma_y * cov_y_x_control)

  # ATE
  
  D = var_t * var_x - cov_t_x^2
  ate = beta1 = (var_x * cov_t_y - cov_t_x * cov_y_x) / D
  beta2 = (var_t * cov_y_x - cov_t_x * cov_t_y) / D
  se_ate <- sqrt((resid_var1 / n_treatment) + (resid_var0 / n_control))
  
  # Relative Effect
  # Correcting the control mean by the covariate imbalance
  mu0_adj <- mean_y_control - gamma_y * (mean_x_control - mean_x)
  rel_effect <- ate / mu0_adj
  
  # SE of Relative Effect
  var_mu0 <- resid_var0 / n_control
  se_rel <- sqrt((se_ate^2 / mu0_adj^2) + (ate^2 * var_mu0 / mu0_adj^4))
  
  return(list(
    ate = ate,
    se_ate = se_ate,
    relative_effect = rel_effect,
    se_relative = se_rel,
    p_value = 2 * (1 - pnorm(abs(ate / se_ate)))
  ))
}
```

# New version

```{r}
compute_ab_stats <- function(n_treatment, sum_y_treatment, sum_x_treatment, sum_y2_treatment, sum_x2_treatment, sum_y_x_treatment,
                             n_control, sum_y_control, sum_x_control, sum_y2_control, sum_x2_control, sum_y_x_control) {

  n <- n_treatment + n_control
  mean_t <- n_treatment / n
  
  # 1. Global Variances and Covariances
  # We use the raw sample variance/cov (1/n) to match OLS matrix math exactly
  var_x <- (sum_x2_treatment + sum_x2_control)/n - ((sum_x_treatment + sum_x_control)/n)^2
  var_t <- mean_t * (1 - mean_t)
  
  cov_tx <- (sum_x_treatment/n) - (mean_t * (sum_x_treatment + sum_x_control)/n)
  cov_ty <- (sum_y_treatment/n) - (mean_t * (sum_y_treatment + sum_y_control)/n)
  cov_xy <- (sum_y_x_treatment + sum_y_x_control)/n - 
            ((sum_x_treatment + sum_x_control)/n * (sum_y_treatment + sum_y_control)/n)

  # 2. Solve the OLS System (y = beta0 + ate*T + slope_x*x)
  # Determinant of the covariance matrix
  D <- var_t * var_x - cov_tx^2
  
  ate <- (var_x * cov_ty - cov_tx * cov_xy) / D
  slope_x <- (var_t * cov_xy - cov_tx * cov_ty) / D

  # 3. Residual Variances (HC0 style)
  # Note: Use slope_x (partial slope), NOT gamma_y (simple slope)
  calc_resid_var <- function(n_g, s_y2, s_x2, s_yx, m_y, m_x, b_x) {
    v_y <- (s_y2 / n_g) - m_y^2
    v_x <- (s_x2 / n_g) - m_x^2
    c_xy <- (s_yx / n_g) - (m_y * m_x)
    # Var(y - slope_x * x) inside the group
    return(v_y + (b_x^2 * v_x) - (2 * b_x * c_xy))
  }

  mean_y_c <- sum_y_control / n_control
  mean_x_c <- sum_x_control / n_control
  mean_y_t <- sum_y_treatment / n_treatment
  mean_x_t <- sum_x_treatment / n_treatment

  v_resid_t <- calc_resid_var(n_treatment, sum_y2_treatment, sum_x2_treatment, sum_y_x_treatment, mean_y_t, mean_x_t, slope_x)
  v_resid_c <- calc_resid_var(n_control, sum_y2_control, sum_x2_control, sum_y_x_control, mean_y_c, mean_x_c, slope_x)

  # 4. Robust SE for ATE
  # This formula accounts for the correlation between T and x (The 1-rho^2 factor)
  rho2 <- (cov_tx^2) / (var_t * var_x)
  se_ate <- sqrt(((v_resid_t / n_treatment) + (v_resid_c / n_control)) / (1 - rho2))

  # 5. Relative Effect (Lift)
  # mu0_adj is the expected y in control if x was at the global mean
  mean_x_global <- (sum_x_treatment + sum_x_control) / n
  mu0_adj <- mean_y_c - slope_x * (mean_x_c - mean_x_global)
  rel_effect <- ate / mu0_adj
  
  # Delta Method for SE of relative lift (assuming mu0_adj is independent of ate for simplicity)
  se_rel <- sqrt((se_ate^2 / mu0_adj^2) + (ate^2 * (v_resid_c/n_control) / mu0_adj^4))

  return(list(
    ate = ate,
    se_ate = se_ate,
    relative_effect = rel_effect,
    se_relative = se_rel,
    p_value = 2 * (1 - pnorm(abs(ate / se_ate)))
  ))
}
```

# Most robust using only second order approximation

```{r}
compute_ab_stats <- function(n_treatment, sum_y_treatment, sum_x_treatment, sum_y2_treatment, sum_x2_treatment, sum_y_x_treatment,
                             n_control, sum_y_control, sum_x_control, sum_y2_control, sum_x2_control, sum_y_x_control) {

  # 1. Global Aggregates
  n <- n_treatment + n_control
  sum_y <- sum_y_treatment + sum_y_control
  sum_x <- sum_x_treatment + sum_x_control
  sum_x2 <- sum_x2_treatment + sum_x2_control
  sum_y_x <- sum_y_x_treatment + sum_y_x_control

  # 2. Build the Global Information Matrix (X'X) 
  # Design: [Intercept, Treatment_Indicator, Covariate]
  XX <- matrix(c(
    n,           n_treatment,     sum_x,
    n_treatment, n_treatment,     sum_x_treatment,
    sum_x,       sum_x_treatment, sum_x2
  ), nrow = 3, byrow = TRUE)

  # X'Y Vector
  XY <- c(sum_y, sum_y_treatment, sum_y_x)

  # Solve for coefficients (beta)
  XX_inv <- solve(XX)
  beta <- XX_inv %*% XY
  
  ate <- beta[2]
  slope_x <- beta[3]

  # 3. Calculate Group-Specific Residual Variances (sigma^2_g)
  # This uses the partial slope (slope_x) to find the unexplained variance
  calc_resid_var <- function(n_g, s_y, s_x, s_y2, s_x2, s_yx, b_x) {
    m_y <- s_y / n_g
    m_x <- s_x / n_g
    v_y <- (s_y2 / n_g) - m_y^2
    v_x <- (s_x2 / n_g) - m_x^2
    cov_yx <- (s_yx / n_g) - (m_y * m_x)
    # Variance of the residuals: Var(y - beta_x * x)
    return(v_y + (b_x^2 * v_x) - (2 * b_x * cov_yx))
  }

  var_resid_t <- calc_resid_var(n_treatment, sum_y_treatment, sum_x_treatment, 
                                sum_y2_treatment, sum_x2_treatment, sum_y_x_treatment, slope_x)
  
  var_resid_c <- calc_resid_var(n_control, sum_y_control, sum_x_control, 
                                sum_y2_control, sum_x2_control, sum_y_x_control, slope_x)

  # 4. Build the "Meat" Matrix (X' * Omega * X)
  # Reconstructing the middle of the sandwich using grouped homoskedasticity
  
  # Treatment Meat Contribution
  XX_t <- matrix(c(
    n_treatment,     n_treatment,     sum_x_treatment,
    n_treatment,     n_treatment,     sum_x_treatment,
    sum_x_treatment, sum_x_treatment, sum_x2_treatment
  ), nrow = 3, byrow = TRUE)
  
  # Control Meat Contribution (Treatment columns/rows are 0)
  XX_c <- matrix(c(
    n_control,       0,               sum_x_control,
    0,               0,               0,
    sum_x_control,   0,               sum_x2_control
  ), nrow = 3, byrow = TRUE)
  
  Meat <- (XX_t * var_resid_t) + (XX_c * var_resid_c)

  # 5. Sandwich Estimator
  V_cov <- XX_inv %*% Meat %*% XX_inv
  se_ate <- sqrt(V_cov[2, 2])

  # 6. Relative Effect (Lift)
  # Adjusting Control Mean by evaluated global mean of x
  mean_x_global <- sum_x / n
  mean_y_c <- sum_y_control / n_control
  mean_x_c <- sum_x_control / n_control
  
  mu0_adj <- mean_y_c - slope_x * (mean_x_c - mean_x_global)
  rel_effect <- ate / mu0_adj
  
  # Delta Method for SE of Relative Effect
  # We use the variance of the adjusted control mean (var_resid_c / n_control)
  var_mu0 <- var_resid_c / n_control
  se_rel <- sqrt((se_ate^2 / mu0_adj^2) + (ate^2 * var_mu0 / mu0_adj^4))

  return(list(
    ate = as.numeric(ate),
    se_ate = as.numeric(se_ate),
    relative_effect = as.numeric(rel_effect),
    se_relative = as.numeric(se_rel),
    p_value = 2 * (1 - pnorm(abs(ate / se_ate)))
  ))
}
```

```{r}
aggs = data %>%
  group_by(treated) %>%
  summarise(
    n = n(),
    sum_y = sum(y),
    sum_y2 = sum(y^2),
    sum_x = sum(x),
    sum_x2 = sum(x^2),
    sum_y_x = sum(y * x)
  )

treatment_aggs = aggs %>%
  filter(treated == 1)
control_aggs = aggs %>%
  filter(treated == 0)

  n_treatment = treatment_aggs %>% pull(n)
  sum_y_treatment = treatment_aggs %>% pull(sum_y)
  sum_x_treatment = treatment_aggs %>% pull(sum_x)
  sum_y2_treatment = treatment_aggs %>% pull(sum_y2)
  sum_x2_treatment = treatment_aggs %>% pull(sum_x2)
  sum_y_x_treatment = treatment_aggs %>% pull(sum_y_x)
  n_control = control_aggs %>% pull(n)
  sum_y_control = control_aggs %>% pull(sum_y)
  sum_x_control = control_aggs %>% pull(sum_x)
  sum_y2_control = control_aggs %>% pull(sum_y2)
  sum_x2_control = control_aggs %>% pull(sum_x2)
  sum_y_x_control = control_aggs %>% pull(sum_y_x)

compute_ab_stats(
  n_treatment = treatment_aggs %>% pull(n),
  sum_y_treatment = treatment_aggs %>% pull(sum_y),
  sum_x_treatment = treatment_aggs %>% pull(sum_x),
  sum_y2_treatment = treatment_aggs %>% pull(sum_y2),
  sum_x2_treatment = treatment_aggs %>% pull(sum_x2),
  sum_y_x_treatment = treatment_aggs %>% pull(sum_y_x),
  n_control = control_aggs %>% pull(n),
  sum_y_control = control_aggs %>% pull(sum_y),
  sum_x_control = control_aggs %>% pull(sum_x) ,
  sum_y2_control = control_aggs %>% pull(sum_y2),
  sum_x2_control = control_aggs %>% pull(sum_x2),
  sum_y_x_control = control_aggs %>% pull(sum_y_x)
)
```