---
title: "Creating CUPED from FWL (under construction)"
author: "Jeffrey Wong"
date: "2026-02-02"
categories: [experimentation, statistics]
---

```{r setup, include=FALSE}
require(dplyr)
require(purrr)
require(ggplot2)
require(gridExtra)
```

[CUPED](https://dl.acm.org/doi/pdf/10.1145/2433396.2433413) is a well known technique to greatly increase statistical power in a controlled experiment. The CUPED formulation has similarities to the OLS model

$$y = \beta_0 + T\beta_1 + (y_{\text{pre}}-1_n \mu_c) \beta_2 + \varepsilon$$

This blog post describes how the CUPED formulation can come out of the
[Frisch Waugh Lovell theorem](https://en.wikipedia.org/wiki/Frisch–Waugh–Lovell_theorem) for
ordinary least squares. This is my favorite way to derive the CUPED estimator,
and this exercise can help to build a deeper appreciation for the algebra of
variance reduction. I particularly like this formulation because the FWL is very
similar to the formulation of other modern causal ML techniques for variance reduction
like the R learner, MLRate, Double ML etc.

In this model, variation in $y$ is explained by the previous values contained in $y_{\text{pre}}$. For example, if someone watched a lot of movies during the pretreatment window, they are likely to watch a lot of movies during the posttreatment window too. Their particular $y$ values are explainable, and we can reduce the residual variance. This increases power.

While there are 3 parameters in the model, only 2 are meaningful. $\hat{\beta}_1$ represents the average treatment effect, and $\hat{\beta}_0$ represents the baseline value for the average user who is in the control, so $\hat{\beta}_0 = E[y | T = 0, y_{\text{pre}} = \mu_c]$. While $\beta_2$ is useful for decreasing variance, it is not useful for inference. It does not play a role in computing the treatment effect. $\beta_2$ is referred to as a nuisance variable.

The CUPED formulation has a special computing strategy, making it not only great for statistical power, but also easy to implement.

# The FWL

If a regression can be partitioned into variables of interest, $T$, and nuisance variables, $X$, then inference on $T$ can be done extremely efficiently.

$$y = \beta_0 + T\beta_1 + X\beta_2 + \varepsilon$$

1. Regress $y$ on $[1, X]$. Then get the residuals, $\tilde{y}$.
2. Regress $T$ on $[1, X]$. Then get the residuals, $\tilde{T}$.
3. Regress $\tilde{y}$ on $\tilde{T}$. The estimated parameters from this regression are exactly $\hat{\beta}_1$ from the original regression.

## Step 1

The regression $y$ on $[1, X]$ is a simple least squares regression with an intercept, e.g. $y = \gamma_{y, 0} + X\gamma_{y, 1} + \nu$. The solution for $\gamma$, and the residuals $\tilde{y}$ are simply

\begin{align}
\hat{\gamma}_{y, 1} &= \frac{Cov(X, y)}{Var(X)} \\
\hat{\gamma}_{y, 0} &= \bar{y} - \hat{\gamma}_{y, 1} \bar{X} \\
\tilde{T} &= y - (\hat{\gamma}_{y, 0} + X \hat{\gamma}_{y, 1})
\end{align}

## Step 2

The regression $T$ on $[1, X]$ can arguably be a no-op. $T$ is by definition not correlated with anything, so the slope of this regression must be 0. The intercept
would just be the proportion of users that are in the treatment. This elegant
subtlety is leveraged in CUPED.

\begin{align}
\hat{\gamma}_{T, 1} &= 0 \\
\hat{\gamma}_{T, 0} &= \bar{T} \\
\tilde{T} &= y - (\bar{T})
\end{align}

However, a fully complete OLS solution would say

\begin{align}
\hat{\gamma}_{T, 1} &= \frac{Cov(X, T)}{Var(X)} \\
\hat{\gamma}_{T, 0} &= \bar{T} - \hat{\gamma}_{T, 1} \bar{X} \\
\tilde{T} &= T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})
\end{align}

## Step 3

Now we regress $\tilde{y}$ on $\tilde{T}$ to get the treatment effect.

$$\hat{\beta} = \frac{Cov(\tilde{y}, \tilde{T})}{Var(\tilde{T})}$$

The numerator can be consolidated.
\begin{align}
Cov(\tilde{y}, \tilde{T}) &= Cov(y - (\hat{\gamma}_{y, 0} + X \hat{\gamma}_{y, 1}), T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})) \\
  &= Cov(y, T) - \hat{\gamma}_{T, 1} Cov(y, X) - \hat{\gamma}_{y, 1} Cov(X, T) + \hat{\gamma}_{y, 1} \hat{\gamma}_{T, 1} Var(X)
\end{align}

The second and fourth terms cancel out because $\hat{\gamma}_{y, 1} = \frac{Cov(X, y)}{Var(X)}$, leaving
$$Cov(\tilde{y}, \tilde{T}) = Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)$$

The denominator is also consolidated
\begin{align}
Var(\tilde{T}) &= Var(T - (\hat{\gamma}_{T, 0} + X \hat{\gamma}_{T, 1})) \\
  &= Var(T) - 2\hat{\gamma}_{T, 1} Cov(T, X) + \hat{\gamma}_{T, 1}^2 Var(X)
\end{align}

Note $\hat{\gamma}_{T, 1} = \frac{Cov(T, X)}{Var(X)}$, which buys the reduction
$$Var(\tilde{T}) = Var(T) - \hat{\gamma}_{T, 1} Cov(T, X)$$
This yields

\begin{align}
\hat{\gamma}_{y, 1} &= \frac{Cov(X, y)}{Var(X)} \\
\hat{\gamma}_{T, 1} &= \frac{Cov(X, T)}{Var(X)} \\
\hat{\beta} &= \frac{Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)}{Var(T) - \hat{\gamma}_{T, 1} Cov(X, T)}
\end{align}

In the same spirit of [conditionally sufficient statistics](https://arxiv.org/pdf/2102.11297),
this compute strategy for the average treatment effect, controlling for a continuous covariate,
can be estimated entirely with aggregated data.

# CUPED is a special case.

CUPED leverages *perfectly* randomized data, so $Cov(X, T) = 0$. Now we will rewrite
the solution for $\hat{\beta}$ in a clever way.

\begin{align}
\hat{\beta} &= \frac{Cov(y, T) - \hat{\gamma}_{y, 1} Cov(X, T)}{Var(T)} \\
  &= \frac{Cov(y, T)}{Var(T)} - \hat{\gamma}_{y, 1} \frac{Cov(X, T)}{Var(T)}
\end{align}
which looks like the difference between 2 different OLS solutions: y regressed on T
and X regressed on T. Knowing T is binary, each of these OLS solutions is simply
the difference in group means:

$$\hat{\beta} = E[y | T = 1] - E[y | T = 0] - \hat{\gamma}_{y, 1} (E[X | T = 1] - E[X | T = 0]).$$
This is exactly the CUPED formulation! However, it truly reveals that CUPED
is like FWL with an assumption that $E[X | T = 1] = E[X | T = 0]$. In small
sample bias this may not be the case; by luck a user with large values for $X$
can be randomized into the treatment. In these scenarios, a pretreatment bias
may be triggered. If it was due to randomization, instead of selection bias, we
would simply control for it by modeling it into the outcome equation. Enforcing
these values to be equal does not let us control for different baselines per arm.

We lean into adopting the full FWL solution, since it also has an efficient compute strategy.
Below, we show how to use it for ratio metrics, as well as doing inference with robust standard errors.

```{r}
# Function to compute treatment effects from aggregated data
compute_ab_stats <- function(n1, sum_y1, sum_p1, sum_y2_1, sum_p2_1, sum_yp1,
                             n0, sum_y0, sum_p0, sum_y2_0, sum_p2_0, sum_yp0) {
  
  # 1. Combined Aggregates
  n <- n1 + n0
  s_y <- sum_y1 + sum_y0
  s_p <- sum_p1 + sum_p0
  s_yp <- sum_yp1 + sum_yp0
  s_p2 <- sum_p2_1 + sum_p2_0
  
  # Means
  y_bar1 <- sum_y1 / n1
  y_bar0 <- sum_y0 / n0
  p_bar1 <- sum_p1 / n1
  p_bar0 <- sum_p0 / n0
  p_bar_total <- s_p / n
  t_bar <- n1 / n
  
  # 2. Covariances (Global)
  var_p <- (s_p2 / n) - (p_bar_total^2)
  cov_yp <- (s_yp / n) - (s_y / n) * (s_p / n)
  cov_tp <- (sum_p1 / n) - (t_bar * p_bar_total)
  cov_ty <- (sum_y1 / n) - (t_bar * (s_y / n))
  var_t <- t_bar * (1 - t_bar)
  
  # 3. FWL Coefficients
  gamma_y <- cov_yp / var_p
  gamma_t <- cov_tp / var_p
  
  # ATE (Beta) - Adjusted for noisy baseline
  ate <- (cov_ty - gamma_y * cov_tp) / (var_t - gamma_t * cov_tp)
  
  # 4. Group-Robust Variances
  # Var(y - gamma*p) = Var(y) + gamma^2*Var(p) - 2*gamma*Cov(y,p)
  var_y1 <- (sum_y2_1 / n1) - (y_bar1^2)
  var_p1 <- (sum_p2_1 / n1) - (p_bar1^2)
  cov_yp1 <- (sum_yp1 / n1) - (y_bar1 * p_bar1)
  resid_var1 <- var_y1 + (gamma_y^2 * var_p1) - (2 * gamma_y * cov_yp1)
  
  var_y0 <- (sum_y2_0 / n0) - (y_bar0^2)
  var_p0 <- (sum_p2_0 / n0) - (p_bar0^2)
  cov_yp0 <- (sum_yp0 / n0) - (y_bar0 * p_bar0)
  resid_var0 <- var_y0 + (gamma_y^2 * var_p0) - (2 * gamma_y * cov_yp0)
  
  # Standard Error of ATE
  se_ate <- sqrt((resid_var1 / n1) + (resid_var0 / n0))
  
  # 5. Relative Effect (Delta Method)
  mu0_adj <- y_bar0 - gamma_y * (p_bar0 - p_bar_total)
  rel_effect <- ate / mu0_adj
  
  # SE of Relative Effect (Delta Method approx)
  # Var(A/B) approx (Var(A)/B^2) + (A^2 * Var(B)/B^4)
  var_mu0 <- resid_var0 / n0
  se_rel <- sqrt((se_ate^2 / mu0_adj^2) + (ate^2 * var_mu0 / mu0_adj^4))
  
  return(list(
    ate = ate,
    se_ate = se_ate,
    relative_effect = rel_effect,
    se_relative = se_rel,
    p_value = 2 * (1 - pnorm(abs(ate / se_ate)))
  ))
}
```